---
title: scATAC-pro Analysis Report
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    social: menu
    theme: united
params:
  output_dir: ~/yuw1/run_scATAC-pro/PBMC10k/output
  configure_user: ~/yuw1/run_scATAC-pro/PBMC10k/configure_user.txt
  
---

<style type="text/css">

.title { /* Header 1 */
  font-size: 28px;
}

.chart-title {  /* chart_title  */
   font-family: "Raleway", Helvetica, Arial, sans-serif;
   font-size: 18px;
   font-weight: 800;
}  
.navbar-brand {
   font-size: 22px;
   font-weight: 900;
}

</style>


<style type="image-container">
  width: 100%;
  margin-left: 0;
  margin-right: 0;
</style>



```{r setup, include=FALSE}
library(flexdashboard)
library(data.table)
library(magrittr)
library(kableExtra)
library(RColorBrewer)
library(ggplot2)
library(plotly)
library(viridis)

```

<span style="font-size:20px;"> Mapping QC </span>
==========================================

Column {data-width=400}
-----------------------------------------------------------------------

### Global mapping statistics

```{r}

read_conf <- function(){
  
  system(paste('grep =', params$configure_user, "|grep -v ^# | awk -F= '{print $1}' | awk '{$1=$1;print}' > vrs.txt "))
  
  system(paste('grep =', params$configure_user, "|grep -v ^# | awk -F= '{print $2}' | awk -F# '{print $1}' | awk '{$1=$1;print}' > vls.txt "))
  
  vrs = readLines('vrs.txt')
  vls = readLines('vls.txt')
  for(i in 1:length(vrs)){
    assign(vrs[i], vls[i], envir = .GlobalEnv)
  }
  system('rm vrs.txt')
  system('rm vls.txt')
}

read_conf()
plotEPS = as.logical(plotEPS)

mapping_qc_file = paste0(params$output_dir, '/summary/', OUTPUT_PREFIX,  '.MappingStats')


mapping_qc = fread(mapping_qc_file, header = F)
mapping_qc$frac = round(mapping_qc$V2/mapping_qc$V2[1], 3)
mapping_qc$frac = paste0(100*mapping_qc$frac, '%')

kable(mapping_qc, col.names = NULL, format = 'html') %>%
  kable_styling("striped", full_width = F, position = 'left')
```


Column {data-width=400}
--------------------------------------------------------
### Mapping statistics for cell barcodes

```{r}
bc_stat_file = paste0(params$output_dir, '/summary/', OUTPUT_PREFIX,  '.qc_per_barcode.txt')
selected_bcs = paste0(params$output_dir, '/filtered_matrix/', CELL_CALLER, '/barcodes.txt')

bc_stat = fread(bc_stat_file)

barcodes = fread(selected_bcs, header = F)$V1

qc_sele = bc_stat[bc %in% barcodes, ]
qc_nonsele = bc_stat[!bc %in% barcodes, ]

fragments_file = paste0(params$output_dir, '/summary/', OUTPUT_PREFIX,  '.fragments.bed')
frags = fread(fragments_file)
names(frags) = c('chr', 'start', 'end', 'bc', 'ndup')

frags = frags[bc %in% barcodes]

ncells = length(barcodes)
frac.in.cell = paste0(round(sum(frags$ndup)/mapping_qc$V2[1], 3) * 100, '%')
med.frag.per.cell = round(median(qc_sele$total_frags))
frac.uniq = paste0(round(nrow(frags)/sum(frags$ndup), 3) * 100, '%')

cell.table = data.frame(c(CELL_CALLER, paste0(ncells), frac.in.cell, paste0(med.frag.per.cell), frac.uniq)) 
rownames(cell.table) = c('Cell called by', 'Estimated # of cells', 'Fraction of MAPQ30 in cells', 
                        'Median fragments per cell', 'Library Complexity (non redudant fraction)')  
kable(cell.table, row.names = T, col.names = NULL, format = 'html') %>%
  kable_styling("striped", full_width = F, position = 'left')


```





<span style="font-size:20px;">Cell barcodes QC  </span> {data-orientation=rows}
==========================================

Row 
--------------------------------------------------------------------
### TSS enrichment score profile
```{r}
tss_escore_file = paste0(params$output_dir, '/signal/', OUTPUT_PREFIX, '.aggregated.mtx.gz')

set.cols = brewer.pal(n=5, name = 'Dark2')
tss.mat = fread(tss_escore_file)
tss.mat = tss.mat[, -c(1:6)]
tss.mat[is.na(tss.mat)] = 0
tss.escore = colSums(tss.mat)
ma <- function(x, n = 10){stats::filter(x, rep(1 / n, n), sides = 2)}
tss.escore = ma(tss.escore)
tss.escore = tss.escore[14:213]
df = data.table(index = 10*(-100:99), escore = tss.escore/tss.escore[1])

p0 <- ggplot(data = df, aes(x = index, y = escore)) + geom_line(size = 1, col = set.cols[1]) +
  xlab('Distance to TSS') + ylab('TSS enrichment score(bp)') + theme(legend.title=element_blank(), axis.text = element_text(size = 15), axis.title = element_text(size = 15))

p0


if(plotEPS){
  system(paste0('mkdir -p ', params$output_dir, '/summary/Figures'))
  pfname = paste0(params$output_dir, '/summary/Figures/tss_enrich.eps')
  ggsave(p0, file = pfname, device = 'eps', width = 6, height = 6)
}


```


### Overlapping with sequence annotated regions
```{r}

qc_sele_df = data.table(frac = c(qc_sele$frac_peak, qc_sele$frac_tss, qc_sele$frac_promoter, qc_sele$frac_enh, qc_sele$frac_mito), 'type' = rep(c('Peaks', 'Tss', 'Promoter', 'Enhancer', 'Mito'), each = nrow(qc_sele)))

p0 <- ggplot(data = qc_sele_df, aes(y = frac, x = type, fill = type)) + ylab('Fraction') +
  geom_boxplot(outlier.size = 0.01) +  theme(legend.position = 'none', axis.text = element_text(size = 15), axis.title.x = element_blank(), axis.title.y = element_text(size = 15)) + xlab('')
#ggplotly(p0)
p0

if(plotEPS){
  pfname = paste0(params$output_dir, '/summary/Figures/overlap_annotation.eps')
  ggsave(p0, file = pfname, device = 'eps', width = 6, height = 6)
}


```




### Overlapping statistics

```{r}

frac_peak = sum(qc_sele$total_frags * qc_sele$frac_peak)/sum(qc_sele$total_frags)
frac_mito = sum(qc_sele$total_frags * qc_sele$frac_mito)/sum(qc_sele$total_frags)
frac_promoter = sum(qc_sele$total_frags * qc_sele$frac_promoter)/sum(qc_sele$total_frags)
frac_enh = sum(qc_sele$total_frags * qc_sele$frac_enhancer)/sum(qc_sele$total_frags)
frac_tss = sum(qc_sele$total_frags * qc_sele$frac_tss)/sum(qc_sele$total_frags)

fracs = data.frame(c(frac_peak, frac_mito, frac_promoter, frac_enh, frac_tss))
row.names(fracs) = c('Fraction in peaks', 'Fraction in mitocondrial chromosome',
                    'Fraction in promoters', 'Fraction in Enhancers(ENCODE)', 
                    'Fraction in TSS')
colnames(fracs) = 'pr'
fracs$pr = round(fracs$pr, 3)
fracs$pr = paste0(100*fracs$pr, '%')

kable(fracs, row.names = T, col.names = NULL) %>%
  kable_styling(full_width = F, position = 'left')



```


Row 
-------------------------------------------------------------------


### Distribution of fragments

```{r}

bc_stat[, 'group' := ifelse(bc %in% barcodes, 'cell', 'non-cell')]

p <- ggplot(data = bc_stat, aes(x = total_frags, col = group)) + 
  geom_density() + scale_x_continuous(trans = 'log10') +
  theme(legend.position=c(.1, .9)) +  theme(legend.title=element_blank(),
                                            axis.title = element_text(size = 15),
                                            axis.text = element_text(size = 15)) + xlab('Total #fragments') + ylab('Density')
p
#ggplotly(p)
```




### Total fragments VS fraction in peaks

```{r}

library(ggplot2)
library(plotly)
g <- ggplot(data = bc_stat[sample(1:nrow(bc_stat), 15000)], aes(x = total_frags, y = frac_peak, col = group)) + geom_point(size = 0.5) + scale_x_continuous(trans='log10') +
      theme(legend.position=c(.9, .9)) +  theme(legend.title=element_blank(),
                                                legend.text = element_text(size=15),
                                                legend.background = NULL,
                                                axis.text = element_text(size = 15),
                                                axis.title = element_text(size = 15)) + xlab('Total #fragments') + ylab('Fraction in Peak')

#ggplotly(g)
g

if(plotEPS){
  pfname = paste0(params$output_dir, '/summary/Figures/fracInPeak_vs_totalFragments.eps')
  ggsave(g, file = pfname, device = 'eps', width = 6, height = 6)
}

```


### Distribution of Insert Size 

```{r}
frags[, 'isize' := end - start]
frags = frags[sample(1:nrow(frags), 1000000), ]

p1 <- ggplot(data = frags[isize < 800], aes(x = isize)) +
  geom_density() + xlab('Insert size') + ylab('Density') +  theme(legend.title=element_blank(), legend.text = element_text(size=15), legend.background = NULL, axis.text = element_text(size = 15), axis.title = element_text(size = 15))

p1

if(plotEPS){
  pfname = paste0(params$output_dir, '/summary/Figures/dist_insert_size.eps')
  ggsave(p1, file = pfname, device = 'eps', width = 6, height = 6)
}


rm(frags)
```


<span style="font-size:20px;"> Downstream Analysis </span>
=====================================================


Column
-----------------------------------------------------

### Clustering 
```{r}
library(Seurat)

down.dir = paste0(params$output_dir, '/downstream_analysis/', CELL_CALLER)
seurat_file = paste0(down.dir, '/seurat_obj.rds')
if(file.exists(seurat_file)){
    ss = readRDS(seurat_file)

    cg <- DimPlot(ss, reduction = 'umap', group.by = 'active_clusters', label = T) + scale_color_brewer(palette = "Paired") + theme(legend.text = element_text(size = 17))
    
    
    
    if(plotEPS){
      pfname = paste0(params$output_dir, '/summary/Figures/umap_clusters.eps')
      ggsave(cg, file = pfname, device = 'eps', width = 6, height = 6)
    }
    
    cg
}


```


### Total reads per cell
```{r}
if(file.exists(seurat_file)){
    ss$log10_nCount = log10(ss$nCount_ATAC)
    #fg <- FeaturePlot(ss, features = 'log10_nCount', reduction = 'umap') + ggtitle('') +
    #   scale_fill_viridis()

    dat = data.table('logN' = ss$log10_nCount, 'UMAP_1' = ss@reductions$umap@cell.embeddings[, 1], 'UMAP_2' = ss@reductions$umap@cell.embeddings[, 2])
ggplot(data = dat, aes(x = UMAP_1, y = UMAP_2, colour = logN)) + geom_point(size = 1.1) +
  scale_colour_viridis() + theme(legend.title=element_blank(), axis.title = element_text(size = 15), axis.text = element_text(size = 15), legend.text = element_text(size = 17))
}

```


Column
----------------------------------------------------

### Motif Analysis

```{r, fig.height=9, out.width='\\textwidth'}
## check enriched TFs for each cluster
library(chromVAR)
library(BiocParallel)
register(SerialParam())

# Do DA/DE with one cluster vs the rest clusters
# clusters are the data frame with <barcode> <cluster>
do_DA <- function(mtx_score, clusters, test = 'wilcox', 
                  only.pos = T, fdr = 0.05, topn = 10){
  clusters$cluster = as.character(clusters$cluster)
  cls = unique(clusters$cluster)
  res = NULL
  features = rownames(mtx_score)
  for(cluster0 in cls){
    bc0 = clusters[cluster == cluster0]$barcode
    mtx1 = mtx_score[, colnames(mtx_score) %in% bc0]
    mtx2 = mtx_score[, !colnames(mtx_score) %in% bc0]
    mu1 = sapply(1:length(features), function(x) mean(mtx1[x, ]))
    mu2 = sapply(1:length(features), function(x) mean(mtx2[x, ]))
    
   
    pvs = sapply(1:length(features), function(x) wilcox.test(mtx1[x, ], mtx2[x, ])$p.value )
    pvs.adj = p.adjust(pvs, method = 'fdr')
    res0 = data.table('feature' = features, 'cluster' = cluster0,
                      'mean1' = mu1, 'mean2' = mu2,
                       'pv' = pvs, 'pv_adjust' = pvs.adj)
    
    if(only.pos) res0 = res0[mean1 > 0]
    
    res0 = res0[order(pv_adjust), ]
    res0 = res0[pv_adjust <= fdr]
    
    if(nrow(res0) > topn) res0 = res0[1:topn, ]
    res = rbind(res, res0)
  }
  return(res)
}


if(file.exists(seurat_file)){
    metaData = ss@meta.data
    rm(ss)
}

if(file.exists(paste0(down.dir, '/chromVar_obj.rds'))){
  chromVar.obj = readRDS(paste0(down.dir, '/chromVar_obj.rds'))
  #variability <- computeVariability(chromVar.obj)
  #variability = data.table(variability, stringsAsFactors = F)
  #variability = variability[order(-variability), ]
  
  ## calculate DA
  da.res = do_DA(chromVar.obj@assays$data$deviations, 
                 clusters = data.table('barcode' = rownames(metaData),
                                       'cluster' = metaData$active_clusters))
  write.csv(da.res, file = paste0(down.dir, '/differential_TF_cluster_enrich.txt'), quote = F, row.names = F )
  
  
  ## plot enriched TFs in heatmap
  sele.tfs = da.res$feature
  zscores = chromVar.obj@assays$data$z
  sele.zscores = zscores[sele.tfs, ]
  
  
  # change tf name to be more readable
  rnames = rownames(sele.zscores)
  nnames = sapply(rnames, function(x) unlist(strsplit(x, '_'))[3])
  nnames1 = sapply(rnames, function(x) unlist(strsplit(x, '_'))[1])
  rownames(sele.zscores) = ifelse(grepl(nnames, pattern = 'LINE'), nnames1, nnames)
  
  metaData$active_clusters = as.character(metaData$active_clusters)  
  metaData = data.table(metaData, keep.rownames = T)
  setkey(metaData, active_clusters)
  
  sele.zscores = sele.zscores[, metaData$rn]
  sele.zscores = sele.zscores[!duplicated(sele.zscores), ]
  
  ann_col = data.table('cluster' = metaData$active_clusters)
  rownames(ann_col) = metaData$rn
  
  sele.zscores[sele.zscores > 10] = 8
  sele.zscores[sele.zscores < -10] = -8
  
  cluster = brewer.pal(n=length(unique(metaData$active_clusters)), name = 'Paired')
  names(cluster) = sort(unique(metaData$active_clusters))
  ann_colors = list('cluster' = cluster)
  
  # resample to reduce memory used
  set.seed(2019)
  rids = sort(sample(1:ncol(sele.zscores), floor(ncol(sele.zscores)/6)))
  ann_col0 = ann_col[rids, ]
  rownames(ann_col0) = colnames(sele.zscores)[rids]
  ph <- pheatmap::pheatmap(sele.zscores[, rids], cluster_cols = F,
                     cluster_rows = T, show_colnames = F, fontsize = 13,
                     annotation_col = ann_col0, color = viridis(100),
                     annotation_colors = ann_colors, fontsize_row = 9)
  
  if(plotEPS){
  pfname = paste0(params$output_dir, '/summary/Figures/heatmap_motif_enrich.eps')
  #postscript(file = pfname, width = 9, height = 12)
  
  ggsave(ph, filename = pfname, device = 'eps', height = 12,
         width = 9)
  #dev.off()
}

}


```

Column
----------------------------------------------------


### Footprinting Analysis: Comparing two clusters

```{r, out.width='\\textwidth'}
comp_cls = paste0('cluster_', cluster1_fp, '_cluster_', cluster2_fp)
footprint_stats.file = paste0(down.dir, '/data_by_cluster/footprint/', comp_cls, '/', comp_cls, '_statistics.txt')
if(file.exists(footprint_stats.file)){
  library(ggrepel)
  library(gridExtra)
#  library(grid)
  footprint_stats = fread(footprint_stats.file)

  footprint_stats[, 'motif1' := unlist(strsplit(Motif, '.', fixed = T))[3], by = Motif]
  footprint_stats[, 'motif2' := unlist(strsplit(Motif, '.', fixed = T))[4], by = Motif]
  footprint_stats[, 'motif2' := ifelse(is.na(motif2), "", motif2)]
  footprint_stats[, 'motif' := paste0(motif1, motif2)]
  footprint_stats[, c('motif1', 'motif2') := NULL]
  
  footprint_stats[, 'isSig' := ifelse(P_values <= 0.05, 'differentiated', 'non-differentiated')]
  footprint_stats[, 'isSig' := ifelse(P_values <= 0.05 & TF_Activity > 0, paste0('cluster', cluster2_fp, '_high'), isSig)]
   footprint_stats[, 'isSig' := ifelse(P_values <= 0.05 & TF_Activity < 0, paste0('cluster', cluster1_fp, '_high'), isSig)]
  footprint_stats$motif_show = ""
  footprint_stats[P_values <= 0.05]$motif_show = footprint_stats[P_values <= 0.05]$motif
  
  
  
 p <- ggplot(data = footprint_stats, aes(x = motif, y = TF_Activity, colour = factor(isSig), label = motif_show)) + geom_point() + xlab("") + ylab('TF Activity Difference') + theme(legend.text = element_text(size=15), axis.title = element_text(size = 15), axis.text.x = element_blank(), legend.title = element_blank(), legend.direction = 'horizontal', plot.title = element_text(size = 15, face = 'bold')) + ggtitle(paste0('TF Activity difference: cluster', cluster2_fp, '-', 'cluster', cluster1_fp  )) + geom_text_repel(force = 10) + theme(plot.margin=unit(c(0.2, 0, 0.2, 0), "cm"), legend.position = 'bottom') + scale_color_manual(values = c('#F8766D', '#619CFF', 'gray40'))
 
 

if(plotEPS){
  pfname = paste0(params$output_dir, '/summary/Figures/footprint_tf_activity.eps')
  
  ggsave(p, filename = pfname, device = 'eps', height = 6,
         width = 6)
  
}
p

}

```


### An example 
```{r, out.width='\\textwidth'}
if(file.exists(footprint_stats.file)){
 
  fdir = dirname(footprint_stats.file)
 #select a TF to show
 
 tf.file.name = footprint_stats[which.min(P_values)]$Motif
 sele.motif = footprint_stats[which.min(P_values)]$motif
 tf.file.name = paste0(fdir, '/Lineplots/', tf.file.name, '.txt')
 
 dd = fread(tf.file.name)
 dd[, 'id' := -99:100]
 dd1 = rbind(subset(dd, select = c('id', paste0('cluster_', cluster1_fp))))
 dd2 = rbind(subset(dd, select = c('id', paste0('cluster_', cluster2_fp))))
 names(dd1)[2] = names(dd2)[2] = 'score'
 dd = rbind(dd2, dd1)
 dd$cluster = rep(c(paste0('cluster_', cluster2_fp), paste0('cluster_', cluster1_fp)), each = nrow(dd1))
 
 
 p2 = ggplot(data = dd, aes(x = id, y = score, colour = cluster)) + geom_line(aes(linetype = cluster), size = 1.1) + ggtitle(sele.motif) + xlab('Distance to Motif') + theme(legend.title=element_blank(), legend.text = element_text(size=15), legend.background = NULL, axis.text = element_text(size = 15), axis.title = element_text(size = 15), legend.position = c(0.8, 0.8), plot.title = element_text(size = 15, face = 'bold')) + theme(plot.margin=unit(c(0.2,0,0.2,0),"cm")) + scale_color_manual(values = c('#F8766D', '#619CFF'))
 
 

 if(plotEPS){
  pfname = paste0(params$output_dir, '/summary/Figures/example_footprint_tf_activity.eps')
  #postscript(file = pfname, width = 9, height = 12)
  
  ggsave(p2, filename = pfname, device = 'eps', height = 6,
         width = 6)
  #dev.off()
}

  p2
}

```
